SSA based global code motion of identical computations

### 2017-04-17 paper.tex
fixes to section 1

Now if the computation of tmax and tmin are hoisted outside the
conditionals, the C code version would look like this:
\begin{verbatim}
float fun(float d, float min, float max, float a) {
  float tmin, tmax;
  float inv = 1.0f / d;
  float x = (min - a) * inv;
  float y = (max - a) * inv;
  if (inv >= 0) { tmin = x; tmax = y; }
  else          { tmin = y; tmax = x; }
  return tmax + tmin;
}

@@

In fact there are several advantages of code-motion of this kind:
\begin{itemize}
\item It helps reduce the code size of the program by replacing multiple
  instructions with one although the final code size may be higher because of
  improved inlining heuristics; functions become cheaper to inline when their
  code size decreases because inliner heuristics in \LLVM{} depends on
  instruction count.
\item It expose more instruction level parallelism to the later compiler
  passes. By hoisting identical computations to be executed earlier, instruction
  schedulers can move heavy computations earlier in order to avoid pipeline
  bubbles;
\item It reduces branch misprediction penalty on out-of-order processors with
  speculative execution of branches: by hoisting/sinking expressions out of
  branches, it can effectively reduce the amount of code to be speculatively
  executed and hence reduce the critical path;
\item It reduces interference/register pressure when appropriate cost-model is
  applied. We have used a cost model in our implementation which hoists/sinks only
  when register pressure would potentially reduce.
\item For SIMD architectures, which execute all branches, it will reduce the
  total number of instructions to be executed.
\item It may improve loop vectorization by reducing a loop with control flow to
  a loop with a single BB, should all the instructions in a conditional get
  hoisted or sinked;
\item It enables more loop invariant code motion (LICM): as LICM passes, in
  general, cannot effectively reason about instructions within conditional
  branches the context of loops, code-hoisting is needed to move instructions
  out of conditional expressions and expose them to LICM.
\end{itemize}

There have been a lot of work both in industry and academia to hoist and sink
code out of branches, and in general global scheduling
\cite{click1995global}. Some relate code-hoisting to code-size optimization
\cite{rosen1988global} and many \cite{barany2013, shobaki2013} use global
scheduling to improve performance. Most of the recent work on global scheduling
are done using ILP which results in prohibitively high compile time. To the best
of our knowledge we have not found any reference which explored code-motion of
identical computation in as much detail as we have done. A part of our
implementation (aggressive code hoisting) is already merged in \LLVM{} trunk,
however, a general implementation of code-motion of congruent (but not
redundant) expressions is still missing from \GCC{} and \LLVM{} trunk. The main
contributions of this paper are:
\begin{itemize}
\item a new optimistic algorithm to move congruent instructiosn out of branches,
\item a cost model to reduce interference and hence, reduce spills,
\item a technique to maximize hoisting in an optimistic approach by partitioning
  the list of potential candidates sorted by their DFS visit number,
\item experimental evaluation of our implementation in \LLVM{} which combines
  \SSA{} based liveness analysis, and ranking of expressions to move very busy
  expressions in order to reduce code-size (and improve performance in some
  cases).
\end{itemize}

@@

\GCC{}ecently got code-hoisting \cite{GCCCodeHoisting} which is implemented as
part of \GVN{}-\PRE{}: it uses the set of ANTIC\_IN and AVAIL\_OUT value
expressions computed for \PRE{}. ANTIC\_IN[B] contains very busy expressions at
basic block B i.e., values computed on all paths from B to exit and
AVAIL\_OUT[B] contains values which are already available. The algorithm hoists
top down to a predecessor.  It uses ANTIC\_IN[B] to know what expressions will
be computed on every path from B to exit, and can be computed in B.  It uses
AVAIL\_OUT[B] to subtract out those values already being computed.  The cost
function is: for each hoist candidate, if all successors of B are dominated by
B, then we know insertion into B will eliminate all the remaining computations.
It then checks to see if at least one successor of B has the value available.
This avoids hoisting it way up the chain to ANTIC\_IN[B].  It also checks to
ensure that B has multiple successors, since hoisting in a straight line is
pointless.  The algorithm continues top down the dominator tree, working in
tandem with \PRE{} until no more hoisting is possible.  One advantage of \GCC{}
implementation is that it works in sync with the \GVN{}-\PRE{} such that when
new hoisting opportunities are created by \GVN{}-\PRE{}, code-hoisting will
hoist them.

>>>>

Now if the computation of tmax and tmin are hoisted outside the
conditionals, the C code version would look like this:
\begin{verbatim}
float fun(float d, float min, float max, float a) {
  float inv = 1.0f / d;
  float x = (min - a) * inv;
  float y = (max - a) * inv;
  float tmin = x, tmax = y;
  if (inv < 0) { tmin = y; tmax = x; }
  return tmax + tmin;
}

@@

In fact there are several advantages of code-motion of this kind:
\begin{itemize}
\item it helps reduce the code size of the program by replacing multiple
  instructions with one although the final code size may be higher because of
  improved inlining heuristics; functions become cheaper to inline when their
  code size decreases because the inliner heuristics depend on instruction
  count;
\item it exposes more instruction level parallelism to the later compiler
  passes. By hoisting identical computations to be executed earlier, instruction
  schedulers can move heavy computations earlier in order to avoid pipeline
  bubbles;
\item it reduces branch misprediction penalty on out-of-order processors with
  speculative execution of branches: by hoisting or sinking expressions out of
  branches, it can effectively reduce the amount of code to be speculatively
  executed and hence reduce the critical path;
\item it reduces interference or register pressure with an appropriate
  cost-model: we have used a cost model in our implementation which only hoists
  or sinks when the register pressure is reduced;
\item it reduces the total number of instructions to be executed for SIMD
  architectures which execute all code in branches based on masking or
  predication;
\item it may improve loop vectorization by reducing a loop with control flow to
  a loop with a single BB, should all the instructions in a conditional get
  hoisted or sinked;
\item it enables more loop invariant code motion (LICM): as LICM passes, in
  general, cannot effectively reason about instructions within conditional
  branches in the context of loops, code-hoisting is needed to move instructions
  out of conditional expressions and expose them to LICM.
\end{itemize}

There have been a lot of work both in industry and academia to hoist and sink
code out of branches, and in general global scheduling
\cite{click1995global}. Some relate code-hoisting to code-size optimization
\cite{rosen1988global} and many \cite{barany2013, shobaki2013} use global
scheduling to improve performance. Most of the recent work on global scheduling
are done using ILP which results in prohibitively high compile time. To the best
of our knowledge we have not found any reference which explored code-motion of
identical computation in as much detail as we have done. A part of our
implementation (aggressive code hoisting) is already merged in \LLVM{} trunk,
however, a general implementation of code-motion of congruent (non redundant)
expressions is still missing from \GCC{} and \LLVM{} trunk. The main
contributions of this paper are:
\begin{itemize}
\item a new optimistic algorithm to move congruent instructiosn out of branches;
\item a cost model to reduce interference and hence, reduce spills;
\item a technique to maximize hoisting in an optimistic approach by partitioning
  the list of potential candidates sorted by their DFS visit number;
\item experimental evaluation of our implementation in \LLVM{} which combines
  \SSA{} based liveness analysis, and ranking of expressions to move very busy
  expressions in order to reduce code-size and improve performance in some
  cases.
\end{itemize}

@@

\GCC{} recently got code-hoisting \cite{GCCCodeHoisting} which is implemented as
part of \GVN{}-\PRE{}: it uses the set of ANTIC\_IN and AVAIL\_OUT value
expressions computed for \PRE{}. ANTIC\_IN[B] contains very busy expressions at
basic block B i.e., values computed on all paths from B to exit and
AVAIL\_OUT[B] contains values which are already available. The algorithm hoists
top down to a predecessor.  It uses ANTIC\_IN[B] to know what expressions will
be computed on every path from B to exit, and can be computed in B.  It uses
AVAIL\_OUT[B] to subtract out those values already being computed.  The cost
function is: for each hoist candidate, if all successors of B are dominated by
B, then we know insertion into B will eliminate all the remaining computations.
It then checks to see if at least one successor of B has the value available.
This avoids hoisting it way up the chain to ANTIC\_IN[B].  It also checks to
ensure that B has multiple successors, since hoisting in a straight line is
pointless.  The algorithm continues top down the dominator tree, working in
tandem with \PRE{} until no more hoisting is possible.  One advantage of \GCC{}
implementation is that it works in sync with the \GVN{}-\PRE{} such that when
new hoisting opportunities are created by \GVN{}-\PRE{}, code-hoisting will
hoist them.
